---
title: "The Null Space of a Matrix"
format: html
---

```{r}
#| echo: false
rm(list=ls())
library(matlib)

A <- matrix(c(-3,6,-1,1,-7,
              1,-2,2,3,-1,
              2,-4,5,8,-4), byrow=TRUE, ncol=5)

rrefA <- matlib::gaussianElimination(A)

ltxA <- matlib::latexMatrix(A, matrix="bmatrix")$matrix
ltxrrefA <- matlib::latexMatrix(rrefA, matrix="bmatrix")$matrix
```
The null space of a matrix, $Null(A)$, is one of the four fundamental subspaces.
The null space is vector space defined by the equation 
$\pmb{A}\pmb{x}=\pmb{0}$. Essentially, the null space is the set of all vectors 
that are mapped to the zero vector by the matrix $\pmb{A}$.

## Finding a basis for $Null(A)$
Consider the following matrix,

$A=`r ltxA`$.

To determine the basis vectors of $Null(A)$, we need to get $pmb{A}$ into 
reduced row echelon form to identify the pivot columns.

$rref(A) = `r ltxrrefA`$

The pivot columns for the reduced row echelon form of $\pmb{A}$ are 1 and 3,
corresponding to the variables $x_1$ and $x_3$.

We call the variables $x_1$ and $x_3$ **basic** variables. They are defined in
terms of the variables $x_2$, $x_4$ and $x_5$. These are known as **free**
variables because they can take on any value. They are not constrained in any
way, hence being dubbed "free".

We can solve for $x_1$ and $x_3$ using the reduced row echelon form. Recall that
$Null(A)$ is defined by $\pmb{A}\pmb{x}=\pmb{0}$ so the first row of $rref(A)$ is
$x_1 - 2x_2 + 0x_3 - x_4 + 3x_5 = 0$. We then solve for $x_1$. We do the exact
same thing for $x_3$ using the second row of $rref(A)$. The resulting equations
give us both $x_1$ and $x_3$ in terms of the free variables.

$$
\begin{aligned}
x_1 &= 2x_2 + x_4 - 3x_5\\
x_3 &= -2x_4 + 2x_5
\end{aligned}
$$

Now, we can rewrite a generic solution of the homogeneous equation 
$\pmb{A}\pmb{x}=\pmb{0}$ as

$$
\begin{bmatrix} x_1 \\ 
x_2 \\ 
x_3 \\ 
x_4 \\ 
x_5 \\ 
\end{bmatrix} = 
\begin{bmatrix} 2x_2 + x_4 - 3x_5 \\ 
x_2 \\ 
-2x_4 + 2x_5 \\ 
x_4 \\ 
x_5 \\ 
\end{bmatrix}
$$
Now, we can decompose the vector on the right hand side into a linear
combination of vectors. This linear combination will form the basis of the null
space of $\pmb{A}$.

$$
\begin{bmatrix} x_1 \\ 
x_2 \\ 
x_3 \\ 
x_4 \\ 
x_5 \\ 
\end{bmatrix} = x_2
\begin{bmatrix} 2 \\ 
1 \\ 
0 \\ 
0 \\ 
0 \\ 
\end{bmatrix} + x_4
\begin{bmatrix} 1 \\ 
0 \\ 
-2 \\ 
1 \\ 
0 \\ 
\end{bmatrix} + x_5
\begin{bmatrix} -3 \\ 
0 \\ 
2 \\ 
0 \\ 
1 \\ 
\end{bmatrix}
$$

Therefore, a basis for $Null(A)$ is
$$
\left\{ 
\begin{bmatrix} 2 \\ 
1 \\ 
0 \\ 
0 \\ 
0 \\ 
\end{bmatrix},
\begin{bmatrix} 1 \\ 
0 \\ 
-2 \\ 
1 \\ 
0 \\ 
\end{bmatrix},
\begin{bmatrix} -3 \\ 
0 \\ 
2 \\ 
0 \\ 
1 \\ 
\end{bmatrix}
\right\}.
$$

## Solutions to an underdetermined system
An underdetermined system is one in which there are fewer equations than there
are variables. This results in a matrix that has fewer rows than columns.

The full solution is a combination of vectors that are mapped to the null space
and those that are mapped into the row space (?) of $\pmb{A}$. Since the system
is underdetermined, there is not a single solution to $\pmb{A}\pmb{x}=\pmb{b}$.

If we let $\pmb{u}$ be a general vector in $Null(A)$ and let $\pmb{v}$ be a 
vector that solves $\pmb{A}\pmb{x}=\pmb{b}$, then $\pmb{x} = \pmb{u} + \pmb{v}$
is a general solution of $\pmb{A}\pmb{x}=\pmb{b}$.

$$
\pmb{A}\pmb{x}=\pmb{A}(\pmb{u}+\pmb{v}) = \pmb{A}\pmb{u}+\pmb{A}\pmb{v} = \pmb{0} + \pmb{b} = \pmb{b}
$$
So, in practice we solve the homogeneous equation and we also find the 
particular solution for a specific $\pmb{b}$. For example, consider the system

$$
\begin{aligned}
2x_1 + 2x_2 + x_3 = 0\\
2x_1 - 2x_2 - x_3 = 1
\end{aligned}
$$

```{r}
#| echo: false
A <- matrix(c(2,2,1,2,-2,-1), byrow=TRUE, ncol=3)
b <- c(0,1)
rref <- matlib::gaussianElimination(A,b,fractions=TRUE)
ltxrref <- matlib::latexMatrix(rref, matrix="bmatrix")$matrix
```

The reduced row echelon form of the augmented matrix is

$rref(A)=`r ltxrref`$.

To solve for the null space basis, we see that we have two basic variables, 
$x_1$ and $x_2$. $x_3$ is free.

Here, $x_1=0$ and $x_2 = -1/2x_3$. Our basis vector for the null space is then
$$
\begin{bmatrix}
0 \\ 
-1/2 x_3 \\ 
x_3
\end{bmatrix} =  x_3
\begin{bmatrix}
0 \\ 
-1/2 \\ 
1
\end{bmatrix}
$$
$x_3$ is arbitrary, the basis vector itself is simply

$$
\begin{bmatrix}
0 \\ 
-1/2 \\ 
1
\end{bmatrix}
$$
Now, we just need to solve for the particular solution. Here, we use the full
augmented matrix.

We can readily see that $x_1=1/4$. Now, since $x_3$ is free, we can pick any 
particular value for it. The most convenient choice is $x_3=0$. Doing this 
results in $x_2 = -1/4$. Therefore, our particular solution is

$$
\begin{bmatrix}
1/4 \\ 
-1/4 \\ 
0
\end{bmatrix}
$$

The complete solution for the system is the combination of these two results.

$$
\begin{bmatrix}
x_1 \\ 
x_2 \\ 
x_3
\end{bmatrix} = a
\begin{bmatrix}
0 \\ 
-1/2 \\ 
1
\end{bmatrix} +
\begin{bmatrix}
1/4 \\ 
-1/4 \\ 
0
\end{bmatrix}
$$

